A formal approach but practical approach in train regulation

### 2025-06-23 paper/main.tex
exemple de regle revu
Avancement règles régulateur

%% Switches
\definecolor{swcolor}{rgb}{0.4, 0.23, 0.56}
\newcommand{\swFmt}[1]{{\color{swcolor}{#1}}}
\newcommand{\sid}[1]{\ensuremath{\swFmt{s_{#1}}}}
\newcommand{\turnouts}{\swFmt{\mathsf{\Theta}}}
\newcommand{\switches}{\ensuremath{\swFmt{\sigma}}}
\newcommand{\networkConf}{\ensuremath{\swFmt{\Sigma}}}
\newcommand{\nosuc}{\ensuremath{\swFmt{\bot}}}

@@

%% Regulator's Program
\newcommand{\regTuple}[3]{\tuple{#1, #2, #3}}
\newcommand{\evGuard}[2]{\ensuremath{\mathtt{ev}(\trainFmt{#1}, #2)}}
\newcommand{\ev}[3]{\evGuard{#1}{#2}: #3}
\newcommand{\incr}[1]{{\mbox{\ensuremath{\mathtt{incr}(\posFmt{#1})}}}\xspace}
\newcommand{\turn}[2]{{\mbox{\ensuremath{\mathtt{turn}(#1, #2)}}}\xspace}
\newcommand{\tokens}{\ensuremath{T}}
\newcommand{\tokenOf}[1]{\ensuremath{T(\posFmt{#1})}}
\newcommand{\regulator}{\ensuremath{R}}

@@

%% Buffer
\newcommand{\bufferFmt}[1]{#1}

@@

\begin{itemize}
	\item The routing program $E$ specifies how the regulator should react to events (i.e. when a train enters a block). A program is a sequence of rules, which are described below.
	\item The token list, denoted $\tokens$, is a function  $\tokens: \blocks \to \Nat$ associating a token level to each block. Those tokens are used to prevent multiple trains accessing the same block at the same time.
	\item The waiting set is a set $W \subseteq \blocks \times \Nat \times \trains$. A triple $\tuple{\bid{id}, n, \tid{id}} \in W$ means block $\bid{id}$ must be protected against train $\tid{id}$ entering as long as $\tokenOf{\bid{id}} < n$.
\end{itemize}

@@

\paragraph{Regulator's rules.}

The syntax of regulator programs is given by the non-terminal {\ttfamily stmt} in Grammar~\ref{gr:regulator_program}.
The program of the regulator is a set of rules, guarded by events. A rule $\ev{\tid{id}}{n}{A}$ denotes that upon entering the $n$-th block of its route, train $\tid{id}$ will trigger actions $A$. There are four kinds of actions: 
\begin{description}
	\item[\incr{\bid{id}}:] this updates the token $\bid{id}$ in $T$:  $T \cdot \incr{\bid{id}}  \triangleq T[\bid{id} \leftarrow T(\bid{id})+1]$
	\item [att(pid, v):] ...\todo{}
	\item [\turn{\sid{id}}{s}:] this instructs the regulator to turn the switch $\sid{id}$ into the $s$ state: $\sigma\cdot\turn{\sid{id}}{s} = \sigma[\sid{id} \leftarrow s]$
	\item [auth:] ...\todo{}
\end{description}

Notice that all actions of a rule are executed atomically. Our semantics, presented below, details actions' behaviour.

\begin{grammar}[Regulator's rules, with $\tid{id} \in \trains$, $n \in  \Nat$, $\bid{id} \in \blocks$, $\sid{id} \in \turnouts$, $s\in \internalState$.][h][gr:regulator_program]
	\firstcasesubtil{\ttfamily stmt}{\evGuard{\tid{id}}{n} : \nonterm{A}^\star}{}
	\firstcase{A}{\incr{\bid{id}}}{}
	\otherform{att(\bid{id}, v)}{}
	\otherform{\turn{\sid{id}}{s}}{}
	\otherform{auth}{}
\end{grammar}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{Start}]{
		\dirFmt{d} \neq \dirFmt{d^\prime}
	}{
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d^\prime}{\trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

Conversely, if the train program is empty (i.e. upon termination), the train stops using rule \ruleRef{Stop}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Stop}]{
		\dirFmt{d} \neq \dirStop
	}{
	\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}} \\
	\reduces
	\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{\dirStop}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{UntilNext}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
	}{
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\push{\bufferFmt{B}}{\tuple{\tid{id}, \bid{id^\prime}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

@@

>>>>

%% Switches
\definecolor{swcolor}{rgb}{0.4, 0.23, 0.56}
\newcommand{\swFmt}[1]{{\color{swcolor}{#1}}}
\newcommand{\sid}[1]{\ensuremath{\swFmt{s_{#1}}}}
\newcommand{\turnouts}{\swFmt{\mathsf{\Theta}}}
\newcommand{\switches}{\ensuremath{\swFmt{\sigma}}}
\newcommand{\networkConf}{\ensuremath{\swFmt{\Sigma}}}
\newcommand{\nosuc}{\ensuremath{\swFmt{\bot}}}
\newcommand{\updateSignals}[3]{\ensuremath{#1[\swFmt{#2}\leftarrow #3]}}

@@

%% Regulator
\newcommand{\handlerOf}[2]{\ensuremath{#1[\trainFmt{#2}]}}
\newcommand{\popHandlerHead}[2]{\ensuremath{{\tt pop}(#1[\trainFmt{#2}])}}
\newcommand{\substHandlerHead}[3]{\ensuremath{#1[\trainFmt{#2} \leftarrow #3]}}

%% Regulator's Program
\newcommand{\regTuple}[3]{\tuple{#1, #2, #3}}
\newcommand{\evGuard}[2]{\ensuremath{\mathtt{ev}(\trainFmt{#1}, #2)}}
\newcommand{\ev}[3]{\evGuard{#1}{#2}: #3}
\newcommand{\incr}[1]{{\mbox{\ensuremath{\mathtt{incr}(\posFmt{#1})}}}\xspace}
\newcommand{\turn}[2]{{\mbox{\ensuremath{\mathtt{turn}(#1, #2)}}}\xspace}
\newcommand{\tokens}{\ensuremath{T}}
\newcommand{\tokenOf}[1]{\ensuremath{T(\posFmt{#1})}}
\newcommand{\regulator}{\ensuremath{R}}
\newcommand{\wait}[2]{{\mbox{\ensuremath{\mathtt{wait}(\posFmt{#1}, #2)}}}\xspace}
\newcommand{\regConcat}[2]{#1 \cdot #2}

@@

%% Buffer
\newcommand{\bufferFmt}[1]{#1}
\newcommand{\head}[1]{\ensuremath{{\tt hd}(#1)}}
\newcommand{\buflen}[1]{\ensuremath{{\tt len}(\bufferFmt{#1})}}
\newcommand{\bufhead}[1]{\head{\bufferFmt{#1}}}
\newcommand{\buftail}[1]{\ensuremath{{\tt tl}(\bufferFmt{#1})}}
\newcommand{\emptyList}{\ensuremath{\varepsilon}}

@@

\begin{itemize}
	\item The routing program $E$ specifies how the regulator should react to events (i.e. when a train enters a block). A program is a collection of rules to apply upon each event (called an \emph{event handler}). We detail further event handlers below.
	\item The token list, denoted $\tokens$, is a function  $\tokens: \blocks \to \Nat$ associating a token level to each block. Those tokens are used to prevent multiple trains accessing the same block at the same time.
	\item The waiting set is a set $W \subseteq \blocks \times \Nat \times \trains$. A triple $\tuple{\bid{id}, n, \tid{id}} \in W$ means block $\bid{id}$ must be protected against train $\tid{id}$ entering as long as $\tokenOf{\bid{id}} < n$.
\end{itemize}

@@

For the sake of clarity $\handlerOf{E}{\tid{id}}$ denotes the list of handlers for events of train $\tid{id}$.

\paragraph{Regulator's event handlers.}

The syntax of regulator programs is given by the non-terminal {\ttfamily stmt} in Grammar~\ref{gr:regulator_program}.
The program of the regulator is a set of rules, guarded by events. A rule $\ev{\tid{id}}{n}{A}$ denotes that upon entering the $n$-th block of its route, train $\tid{id}$ will trigger actions $A$. The list of actions in the handler are of four kinds: 
\begin{description}
	\item[\incr{\bid{id}}:] this updates the token $\bid{id}$ in $T$:  $T \cdot \incr{\bid{id}}  \triangleq T[\bid{id} \leftarrow T(\bid{id})+1]$
	\item [\wait{\bid{id}}{i}:] ...\todo{}
	\item [\turn{\sid{id}}{s}:] this instructs the regulator to turn the switch $\sid{id}$ into the $s$ state: $\sigma\cdot\turn{\sid{id}}{s} = \sigma[\sid{id} \leftarrow s]$
	\item [auth:] ...\todo{}
\end{description}

Notice that all actions of a rule are executed atomically. Our semantics, presented below, details actions' behaviour.
\todo{Clarifier Event, indexation par train, qui donne une liste de handlers, eux-même une liste d'ordres}

\begin{grammar}[Regulator's rules, with $\tid{id} \in \trains$, $n \in  \Nat$, $\bid{id} \in \blocks$, $\sid{id} \in \turnouts$, $s\in \internalState$.][h][gr:regulator_program]
	\firstcasesubtil{\ttfamily stmt}{\evGuard{\tid{id}}{n} : \nonterm{A}^\star}{}
	\firstcase{A}{\incr{\bid{id}}}{}
	\otherform{\wait{\bid{id}}{i}}{}
	\otherform{\turn{\sid{id}}{s}}{}
	\otherform{auth}{}
\end{grammar}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{Start}]{
		\dirFmt{d} \neq \dirFmt{d^\prime} \and P = \trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}
	}{
		\hspace{3mm} \redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{P}\}}{\regulator}{\switches}{\signals}} \\
		\reduces
		\redTuple{\guardT}{\push{B}{\tuple{\tid{id}, \bid{id}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d^\prime}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

Conversely, if the train program is empty (i.e. upon termination), the train stops using rule \ruleRef{Stop}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Stop}]{
		\dirFmt{d} \neq \dirStop
	}{
	\hspace{4mm}  \redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}} \\
	\reduces
	\redTuple{\guardT}{\push{B}{\tuple{\tid{id}, \bid{id}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{\dirStop}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{UntilNext}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
	}{
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\push{\bufferFmt{B}}{\tuple{\tid{id}, \bid{id^\prime}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}\todo{Pas de message dans ce cas là?}

@@

When the regulator reacts to a train event, it executes atomically all rules associated with that event. To implement this atomic execution, the first regulator event is to change the guard to \guardR (see Fig. \ref{fig:state_machine_guard}).

\begin{mathpar}
	\inferrule*[left=\ruleDef{StartEvent}] {
		\buflen{B} \neq 0
	}{
		\tuple{\guardT, B, R} \reduces \tuple{\guardR, B, R}
	}
\end{mathpar}

Then, the regulator lookup the first event on the buffer, and executes all the rules in the event handler associated with this event. All actions are treated in order, and when the list of remaining actions is empty, the handling terminates and the event is removed of the buffer. We introduce 6 reduction rules for the 4 possible kinds of orders (\incr{\bid{id}} and \wait{\bid{id}}{i} have two rules).

Rule \ruleRef{Turn} triggers when a \turn{\sid{id}}{s} is encountered in the event handler. It updates the state of the targeted switch.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Turn}]{
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \regConcat{\turn{\sid{id}}{s}}{A^\prime}
	}{
	\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
	\reduces
	\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{\substHandlerHead{E}{\tid{id}}{A^\prime}, T, W}}{\updateSignals{\switches}{\sid{id}}{s}}{\signals}}
	}
\end{mathpar}

\todo{other rules}

Finally, when the handler of the pending event has no action leftover, the regulator terminates its handling: it removes the (now empty) head of $\handlerOf{E}{\tid{id}}$, the head of the buffer, and returns the guard to state $\guardT$.

\begin{mathpar}
	\inferrule*[left=\ruleDef{EndEvent}]{
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \emptyList
	}{
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
		\reduces
		\redTuple{\guardT}{\buftail{B}}{\stateTuple{\trainSeq}{\tuple{\popHandlerHead{E}{\tid{id}}, T, W}}{\switches}{\signals}}
	}
\end{mathpar}
